\section{AI Layer}
\label{sec:ai}

\subsection{Overview}
\label{sec:ai-overview}

The AI layer sits on top of the Knowledge Graph and provides both the user and
the system with access to an AI model that has structured context about what is
happening on the machine. It is not a chatbot grafted onto an existing OS; it
is an interface that can query the graph, invoke application capabilities
through a standardized protocol, and optionally act autonomously when the user
explicitly enables that.

The AI layer is not the primary reason this system exists. The desktop, the
permission system, the profile isolation, and the store all stand on their own.
AI is an enhancement that becomes more useful as the Knowledge Graph fills in
over time, and users who want none of it can disable it completely without
losing any other functionality.

Three principles govern the design of this layer.

\paragraph{Query-first.}
The AI responds when asked. Nothing happens in the background without an
explicit trigger. Autonomous features exist but are separate opt-ins, not
defaults that the user has to find and disable.

\paragraph{Provider-agnostic.}
The system does not care whether the model runs locally or in the cloud.
Every provider implements the same interface; the rest of the codebase never
speaks to a specific backend. The user decides what runs where, including
routing sensitive requests to a local model only.

\paragraph{Opt-out at any point.}
The entire layer can be disabled in Settings under AI. When disabled, no local
model runs, no graph queries originate from the AI layer, Spotlight falls back
to standard search, and no MCP servers are active. The setting is not hidden
five menus deep.

\subsection{Cold-Start Behavior}
\label{sec:ai-coldstart}

The Knowledge Graph is empty on a fresh install. An AI feature that promises
contextual awareness but has nothing in the graph yet is worse than no feature
at all: it creates an expectation and immediately fails to meet it. The system
handles this with straightforward communication rather than by faking
usefulness.

During onboarding, AI features are presented with an honest explanation of
what to expect: the assistant learns context gradually as the system is used,
starts with nothing, and becomes more useful over days and weeks. There are no
demo interactions, no placeholder context, and no fabricated examples of what
it will eventually be able to do.

At onboarding the user can optionally seed the graph with existing data from
their current system: browser history, calendar events, and documents. This is
presented as a trade-off, not a recommendation. Each source can be imported
independently, so someone who is comfortable sharing their calendar but not
their browsing history can import only the former. Importing gives the AI
immediately useful context without a multi-week waiting period.

The AI assistant does not surface proactively until a threshold is met: seven
days of usage or a completed manual import. Before that threshold the feature
is accessible but does not advertise itself. The non-AI parts of the system are
functional and complete from the first boot, so there is no gap in usefulness
during this period.

\subsection{Provider Abstraction}
\label{sec:ai-providers}

Different AI providers have different APIs, different latency profiles, and
significantly different privacy implications. Sending the contents of a private
file to a cloud API is appropriate in some contexts and completely unacceptable
in others. The system needs to handle both without the rest of the codebase
caring which provider is active.

The solution is a Rust trait that every AI backend implements. The AI layer
constructs a \texttt{CompletionRequest}, passes it to whichever provider is
configured, and receives a \texttt{CompletionResponse}. No provider-specific
code is visible outside the provider module.

\begin{lstlisting}[language=Rust, caption={AI provider trait}]
#[async_trait]
pub trait AiProvider: Send + Sync {
    async fn complete(
        &self,
        request: CompletionRequest,
    ) -> Result<CompletionResponse>;
    async fn available(&self) -> bool;
    fn name(&self) -> &str;
}
\end{lstlisting}

Figure~\ref{fig:providers} shows how the AI layer relates to the available
backends.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  box/.style={
    draw, thick, rectangle, rounded corners=2pt,
    minimum width=3.2cm, minimum height=0.8cm,
    font=\small, align=center, inner sep=5pt
  },
  arr/.style={->, thick},
]

\node[box] (ail)   at (0, 3.2) {\textbf{AI Interface Layer}};
\node[box] (trait) at (0, 2.0) {Provider Trait (Rust)};
\node[box] (local) at (-2.4, 0.8) {Local Provider};
\node[box] (cloud) at ( 2.4, 0.8) {Cloud Provider};
\node[box, dashed] (lm)  at (-2.4, -0.4) {\footnotesize Ollama / llama.cpp};
\node[box, dashed] (api) at ( 2.4, -0.4) {\footnotesize OpenAI / Anthropic / \ldots};

\draw[arr] (ail)   -- (trait);
\draw[arr] (trait) -- (local);
\draw[arr] (trait) -- (cloud);
\draw[arr] (local) -- (lm);
\draw[arr] (cloud) -- (api);

\end{tikzpicture}
\caption{Provider abstraction. The AI layer speaks only to the trait; the
concrete backend is a configuration choice. Local and cloud providers are
interchangeable from the perspective of anything above the trait boundary.}
\label{fig:providers}
\end{figure}

Multiple providers can be active simultaneously under a routing policy
configured by the user: a local model as the default, with fallback to cloud
if the local model is unavailable or too slow; specific task types such as code
always routed to a designated provider; and sensitive contexts, such as
requests involving personal file contents, explicitly prevented from leaving
the machine.

\subsection{MCP Integration}
\label{sec:mcp}

\emph{MCP} (Model Context Protocol) is an open protocol that standardizes how
AI models interact with external tools and data sources. Before MCP, every AI
integration was bespoke: one application called the OpenAI API directly,
another used its own tool-calling format, and combining them required custom
glue code on both ends. MCP standardizes this at the protocol level so that
any MCP-compatible AI client can communicate with any MCP server without
application-specific integration work.

An MCP server is a small process that exposes a set of tools the AI can call
and resources the AI can read. In this system, first-party applications expose
MCP servers as part of their standard implementation: the file manager exposes
\texttt{list\_directory}, \texttt{open\_file}, and \texttt{move\_file}; the
terminal exposes \texttt{run\_command} and \texttt{get\_output}; the Knowledge
Graph exposes a query interface. The AI can coordinate across multiple
applications in a single interaction by calling their respective servers in
sequence. Figure~\ref{fig:mcp} shows the topology.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  box/.style={
    draw, thick, rectangle, rounded corners=2pt,
    minimum width=2.8cm, minimum height=0.8cm,
    font=\small, align=center, inner sep=5pt
  },
  srv/.style={
    draw, thick, rectangle, rounded corners=2pt,
    minimum width=2.8cm, minimum height=0.8cm,
    font=\small, align=center, inner sep=5pt
  },
  arr/.style={->, thick},
]

\node[box] (model)  at (0, 0)   {\textbf{AI Model}};
\node[box] (client) at (3.4, 0) {MCP Client};
\node[srv] (fm)     at (6.8,  1.8) {MCP: File Manager};
\node[srv] (term)   at (6.8,  0.6) {MCP: Terminal};
\node[srv] (kg)     at (6.8, -0.6) {MCP: Knowledge Graph};
\node[srv] (ext)    at (6.8, -1.8) {MCP: third-party};

\draw[arr] (model)  -- (client);
\draw[arr] (client) -- (fm);
\draw[arr] (client) -- (term);
\draw[arr] (client) -- (kg);
\draw[arr] (client.east) -- (ext);

\end{tikzpicture}
\caption{MCP topology. The AI model communicates through a single MCP client
that dispatches to individual application servers. Third-party applications
from the Store can ship their own MCP servers, subject to the permission model
described in Section~\ref{sec:mcp-permissions}.}
\label{fig:mcp}
\end{figure}

CLI tools do not need MCP servers. If a tool has a \texttt{-{}-help} flag and
produces sensible output, the AI can invoke it as a subprocess and read
\texttt{stdout}. MCP is reserved for GUI applications and stateful interfaces
where a subprocess call is insufficient.

\subsection{Interaction Model}
\label{sec:ai-interaction}

\paragraph{Phase 1: Query-based (default).}
The user asks a question, the AI responds. No background activity occurs
without an explicit trigger. The AI has access to the Knowledge Graph and
whatever MCP servers are permitted under the current configuration. Typical
interactions include querying session history, finding files related to a
project, correlating system events with observed behaviour, and composing
multi-step actions across applications.

When a question is asked, the AI layer translates the natural language input
into a Cypher query, executes it against the Knowledge Graph via the Graph
Daemon, and passes the structured results to the model as context. Relative
time expressions are resolved to absolute UTC timestamps before the generation
step, so the model always operates on concrete values rather than ambiguous
phrases. The Cypher generation and the result formatting are two separate model
calls, since combining them causes the model to bias the query toward a
desired-sounding answer rather than accurately querying the schema.

The query validator sits between generation and execution: it checks that the
generated Cypher uses only known node types and relations, contains no write
operations, and includes a \texttt{LIMIT} clause. If validation fails, the
error is fed back for a retry; after two failed attempts the user is told the
question could not be answered, with no silent fallback.

\paragraph{Phase 2: Autonomous agent (explicit opt-in).}
A background agent that observes the event stream and acts proactively is
available as an opt-in capability. It is not enabled by default and each
autonomous behaviour is a separate toggle: suggesting related files when a
document is opened is independent of automatically tagging new files by
project. The agent uses the same provider abstraction and MCP layer as the
query interface; the difference is that it initiates actions rather than
responding to them.

\subsection{MCP Permission Model}
\label{sec:mcp-permissions}

MCP servers are the interface through which the AI acts on applications.
The permission model distinguishes between reading information, which the AI
can do by default within the user's configured AI permission level, and
modifying state, which requires explicit per-session authorization.

\begin{table}[htbp]
\caption{Read-only MCP servers (permitted by default)}
\label{tab:mcp-read}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Server} & \textbf{Exposes} \\
\midrule
File Manager   & Directory listings, file metadata \\
Calendar       & Upcoming and past events \\
Browser        & History and open tabs (no passwords, no form data) \\
System Monitor & Process list and resource usage \\
Notes          & Note titles and content \\
\bottomrule
\end{tabularx}
\end{table}

Reading to answer a question is what the user expects when they ask the AI
something. Answering "what meetings do I have today?" without a confirmation
dialog is the correct behaviour. Reading does not change state, and the AI
permission level configured in Settings still gates what data each server
exposes regardless of this default.

\begin{table}[htbp]
\caption{Write and action MCP servers (per-session authorization required)}
\label{tab:mcp-write}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Server} & \textbf{Can do} \\
\midrule
Terminal     & Execute shell commands \\
File Manager & Create, move, rename, delete files \\
Calendar     & Create and modify events \\
Email        & Compose and send messages \\
Store        & Install and remove packages \\
\bottomrule
\end{tabularx}
\end{table}

Authorization is granted once per session and covers the scope the user
specifies: approving file management does not imply approval for email. It
expires at session end and is not persisted. Within an authorized scope, a
separate hardcoded list of actions always requires explicit confirmation
regardless of session permissions, because their consequences are irreversible
or high-impact: permanent file deletion, sending any external message,
installing or removing system packages, modifying system configuration outside
\texttt{\textasciitilde/.config}, any network request to a host outside the
application's declared permissions, and running commands with elevated
privileges. This list is not configurable.

Third-party MCP servers shipped by Store applications are always treated as
action-requiring by default. A community server that claims to be read-only
receives the same treatment until it carries a verified security audit covering
its MCP implementation.
